---
title: "Lecture 2"
description: "Getting Started with R"
date: "08-31-2026"
format: html
editor: visual
---

# Week 2

## Lecture 1

### Setting the working directory

You can set the working directory using this command:

```{r}
#setwd(...) 
```

You can see the working directory using this command:

```{r}
getwd()
```

### Review of Data Types

An $r \times c$ **matrix** is a rectangular array of symbols arranged in $r$ rows and $c$ columns.

$$A = \begin{bmatrix} 1 & 2 \\ 6 & 3 \end{bmatrix}$$

A **column vector** is an $r \times 1$ matrix, a matrix with only one column.

$$A = \begin{bmatrix} 1 \\ 6
 \end{bmatrix}$$

A **row vector** is a $1 \times c$ matrix, a matrix with only one row.

$$A = \begin{bmatrix} 7&12
 \end{bmatrix}$$

A $1 \times 1$ matrix is called a scalar, but it's just an ordinary number.

$$A = [4]$$

To define a vector in R with the numbers `3, 4, 5`, we can use the function `c()` which is short for `concatenate`:

```{r}
b = c(3,4,5)
b
```

### Functions

There are many different kinds of functions in R:

```{r}
mean(b)
sum(b)
```

The following code generates and plots 10 random numbers from the standard normal distribution:

```{r}
x <- round(rnorm(10, mean=0, sd=1), 3)
plot(x)
```

### Vectors

```{r}
vec1 = c(1,2,3,4,5,6)
vec1

vec1[5]

vec1[3] = 12 
vec1

vec2 = seq(from=0, to=1, by=0.25)
vec2

sum(vec1)
vec1 + vec2
```

### Matrices

Matrices are like 2-dimensional vectors (or like data tables)

```{r}
mat = matrix(data=c(9,2,3,4,5,6), ncol=3)
mat
```

The argument `data` specifies which numbers should be in the matrix. We use either `ncol` to specify the number of column or `nrow` to specify the number of rows.

Matrix-operations are similar to vector operations:

```{r}
mat[1,2]

r2 <- mat[2,]
r2

c2 <- mat[,2]
c2

mean(mat)
```

```{r}
mat1 = matrix(data=c(9,2,3,4,5,6), ncol=3)
mat1

mat2 = matrix(data=c(7,6,5,3,2,8), nrow=2)
mat2

matsum=mat1+mat2
matsum
```

In order to append 2+ matrices, the number of columns has to be the same; in order to merge, the number of rows has to be the same.

```{r}
merging = cbind(mat1, mat2)
merging
```

```{r}
appending = rbind(mat1,mat2)
appending
```

```{r}
trans=t(appending)
trans

dim(trans)
```

### Arrays

Arrays are like matrices, but they have 2 or more dimensions. Think of them as an **array of arrays**.

```{r}
z <- array(1:24, dim=c(2,3,4))
z
```

### Data Frames

A data frame is a matrix with names above the columns. You can call and use one of the columns without knowing which position it is:

```{r}
t = data.frame(x=c(11,12,14), y=c(19,20,21), z=c(10,9,7))
t
```

```{r}
mean(t$z) # or mean(t[['z']])
```

### Lists

Unlike matrices and data frames, the "columns" of lists don't have to be the same length:

```{r}
L = list(one=1, two=c(1,2), five=seq(0,1,length=5))
L
names(L)
L$five + 10
```

### Reading and writing files

```{r}
d = data.frame(a=c(3,4,5), b=c(12,43,54))
d
```

```{r}
write.table(d, file='tst0.txt', row.names=FALSE)
```

```{r}
d2 = read.table(file='tst0.txt', header=TRUE)
d2
```

### Reading and writing shapefiles

```{r}
#install.packages('sf')
#install.packages('ggplot2')
library(sf)
library(ggplot2)
```

```{r}
PhillyData <- st_read('/Users/michaelpincus/Documents/MUSA/Statistics/statistics/RegressionData/Regression Data.shp')
```

```{r}
plot(st_geometry(PhillyData), border='red')
```

A better way to plot it is with `ggplot2`:

```{r}
ggplot() + 
  geom_sf(data=PhillyData, aes(fill=MEDHHINC), color='white', linewidth=0.01) + 
  labs(
    title = 'Philadelphia 2000 Data',
    subtitle = 'Census BGs',
    fill = 'Median Household Income'
  ) + theme_void()
```

### Missing Data

We have some missing data here for variables `a` and `c`:

```{r}
NAEx = read.csv('/Users/michaelpincus/Documents/MUSA/Statistics/statistics/Week2/Missing Data.csv')
NAEx
```

```{r}
max(NAEx)
max(NAEx, na.rm=TRUE)
```

The `na.rm` command tells R whether `NA` values should be stripped before the computation proceeds.

```{r}
NAEx$d = NAEx$c + 1
NAEx
```

### Strings

Sometimes we want to specify something which is not a number, ex. the name of a measurement station or a data file.

```{r}
m = "apples"
m

n = "pears"
n
```

### If-Else Statements

```{r}
w = 3 
if(w<5) {d=2} else {d=10}
d
```

### Functions

```{r}
fun1 = function(arg1,arg2)
{
  w=arg1^2
  return(arg2+w)
}

fun1(arg1=3, arg2=5)
```

Some useful statistical functions:

```{r}
a = c(1,1,1,1,1,1,0,0,0,0,0,0,0,2,2,2,3,3,3,4,4)

mean(a)

max(a)

round(sd(a),2)

sum(a)

summary(a)

summary(factor(a))
```

The variable $r$ test
