---
title: "Lecture 6"
description: "Multiple Regression and Outputs"
date: 09-16-2026
editor: visual
---

## Review from Lecture 5

Assumptions:

- Residuals are normal: $\epsilon \sim \mathcal{N}(\mu=0, \sigma^2)$

- Residuals are random

We are trying to minimize the $\text{SSE}$, the sum of squared errors, when we are creating a regression equation.

$R^2$ is the total variance in data as explained by the model. It is obtained through this formula:

$$
R^2 = \frac{\text{SSR}}{\text{SST}}
$$

Where $\text{SSR}$ is the sum of squared residuals (variance explained by the model) and $\text{SST}$ is the total sum of squares (the variance in the actual data). Note that:

$$
\text{SST} = \text{SSR + SSE}
$$

Where $\text{SSE}$ is the sum of squared errors, the unexplained variance in residuals.

### In other words

$\text{SST}$ (Total Sum of Squares) doesn't involve your model at all, it's just how much $y$ values are spread around the average.

$\text{SSR}$ (Regression Sum of Squares) tells you how far your predicted values are from the mean. It measures the variation that the model's prediction captures.

$\text{SSE}$ (Error Sum of Squares) measures the variation your model fails to explain.

## Multiple Regression

When we have $k>1$ predictors, rather than getting a line in 2 dimensions, we get a surface in $k+1$ dimensions

Each independent variable will have its own slope coefficient which will indicate the relationship of that particular predictor with the dependent variable, *controlling for all other independent variables in the regression*.

The best fit equation becomes:

$$
\text{Income} = \beta_0 + \beta_1x_1 + \beta_2x_2+\dots+\beta_kx_k+\epsilon
$$

The coefficient $\beta_i$ of each predictor may be interpreted as the amount by which the dependent variable changes as the independent variable increases by one unit (holding all other variables constant)

### $\beta$ coefficients

$\beta$ is the coefficient of the independent variable (i.e., the slope of the regression line)

- There is one coefficient per independent variable

- The sign indicates whether the relationship between the dependent and independent variables is positive (direct) or negative (inverse)

- The value of $\beta$ is the amount by which the dependent variable changes when the predictor increases by one unit (all other predictors being held constant)

### $\beta$ coefficient estimation - least squares

Given $n$ observations on $y$, and $k$ predictors $x_1 \dots x_k$, the estimates $\hat{\beta_0}, \hat{\beta_1}, \hat{\beta_2}, \dots, \hat{\beta_k}$ are chosen simultaneously to minimize the expression for the Error Sum of Squares $\text{SSE}$ given by:

$$
\text{SSE}=\sum_{i=1}^n\epsilon^2 = \sum_{i=1}^n(y-\hat{y})^2
$$

### $R^2$ and adjusted $R^2$

As in simple regression, $\text{SST}=\sum(y_i-\bar{y})^2$ and $R^2=1-\frac{\text{SSE}}{\text{SST}}$.

Here, $R^2$ is the coefficient of multiple determination, or the proportion of variance in the model explained by all $k$ predictors.

Because extra predictors will generally increase $R^2$, it is typically adjusted for the number of predictors:

$$
R^2_{\text{adj}}=\frac{(n-1)R^2-k}{n-(k+1)}
$$

### Model Utility Test: $F$-test (or the $F$-ratio)

Before doing a hypothesis test for each individual predictor, we do a so-called model utility test, referred to as the $F$-test. It is a goodness of fit measure, as is $R^2$

It tests $H_0$ that all coefficients in the model are (jointly) $0$ vs. $H_1$ that at least one of the coefficients $\ne 0$

$$
H_0: \beta_1=\beta_2=\dots=\beta_k=0
$$

$$
H_1: \beta_1\ne0 \text{ or } \beta_2 \ne 0\text{ or } \dots \ \beta_k\ne0
$$

## Hypothesis testing and confidence intervals

In addition to the omnibus utility model the $F$-test, a hypothesis test (the $t$-test) is done for each predictor $i$.

Most statisticians will only look at these $t$-tests for each individual predictor if the $F$-test is significant.

As in simple regression, $H_0$ is always that the predictor $i$ is not associated with the dependent variable (i.e., $H_1 : \beta_i=0$ ).

The alternative hypothesis is that $\beta_i\ne0$.

### $p$-values

The $p$-value is the probability of observing a value that is at least as different from $0$ (the value stated in $H_0$ as the given estimated value.

- If this probability is small enough (generally $p<0.05$), we reject $H_0$ of $\beta=0$ for an $H_1$ of $\beta \ne 0$.

- The rejection of $H_0$ indicates that the independent variable is a statistically significant predictor of the dependent variable.

- Again, when the null hypothesis $\beta=0$ cannot be rejected, the dependent variable is not related to the independent variable.

- There is one $p$-value per independent variable.

## Multiple regression assumptions

1.  Same assumptions and model aptness checks as for simple regressions
2.  The predictors should not be very strongly correlated with each other (i.e., no *multicollinearity*)
    - Multicollinearity is when two or more predictors are very strongly correlated with eachother, so that $r>0.9$ or $r<-0.9$.

### Better ways of testing multicollinearity

Imagine you have a regression with $k$ predictors.

If $k>2$, it is possible that the pairwise correlations are small, and yet linear dependence exists among three or even more variables. Sometimes there exists a situation where:

$$
X_3 \approx X_1 + X_2
$$

In these instances, you can assess for multicollinearity by regressing each predictor $k$ on all the remaining predictors. Basically, you treat it temporarily as if it were the outcome and see how well the **other** predictors can predict it.

- Look at the $R^2$ in each of the $k$ regressions. If any of these regressions yield $R^2 > 0.8$, there is likely a problem with multicollinearity.

- Using the $R^2$ values from above, calculate the Variance Inflation Factor for each predictor $k$ in the *original* regression. You can do this with `vif()` in R. By hand $\text{VIF}$ is calculated as:

$$
\text{VIF}_k=\frac{1}{1-R^2_k}
$$

We get one $\text{VIF}$ for each predictor in our original regression. The general rule of thumb is that if $\text{VIF}>4$, it warrants further investigation.

### Number of Observations 

A major rule of thumb is that you should have *at least* 10 observations for every parameter estimated. So with $n=15$, you should not include more than one $\beta$ in your regression.

## Regression outputs 

### One continuous predictor 

Let's say we are running a regression on aggravated assaults using median house value as a predictor and get the following results:

- `R-SQUARE: 0.23`

- `CONSTANT: 34.32, Probability: 0.0000000`

- `MEDHVALUE: -0.91, Probability: 0.0000000`

23% of the variance in our dependent variable is explained by the model `R-SQUARE`. Median house value is a highly significant predictor of aggravated assaults (`p<0.000001`). A 1 unit increase in the median house value is associated with a decrease in the number of aggravated assaults by `0.91`.

### Two continuous predictors 

Let's say we want to include the presence of parks in the predictor block group (a binary predictor, `0` or `1`. We get the following results:

- `R-SQUARE: 0.27, ADJUSTED R-SQUARE 0.26`

- `CONSTANT 35.42, Probability: 0.000000`

- `MEDHVALUE -0.86, Probability: 0.000000`

- `PARKS 8.31, Probability: 0.000000`

27% of the variance in our dependent variable is explained by the model. Both median house value and the presence of parks are highly significant predictors of aggravated assaults (`p<00000.1` for both variables). 1 unit decrease in median house value is associated with a decrease in the number of aggravated assaults by `0.86`. A one-unit increase int he presence of parks is associated with `8.31` additional aggravated assaults.

It is important to note that adding new predictors changes the $\beta$ coefficients and corresponding $p$-values of the variables that were originally in the model.

## Categorical/Ordinal variables as predictors 

Assume you'd like to add another predictor to the model: the community policing organization, or unit, serving the block group

- Assume that the community policing unit variable has the following values: A, B, C, D

- Recall that $\beta$ is the amount by which the dependent variable changes should the predictor go up by 1. When we have a categorical predictor, the interpretation of $\beta$ becomes nonsensical.

- The way to include categorical and ordinal variables is to create a set of **dummy** (binary) variables

If you have four values in the community policing unit, any three dummy variables will be included in the model. With this method, the first three dummy variables take on `1` if they are the unit included in the block group, and `0` if otherwise. If all three are `0`, then it's bound to be unit D. The reason why we can't include all four in the model:

$$
\text{DummyA + DummyB + DummyC + DummyD}=1
$$

### Practice with dummy variables in R

```{r}
set.seed(123)
n <- 150

df <- data.frame(
  Department = sample(c("HR", "Marketing", "Engineering"), n, replace = TRUE),
  Years      = round(runif(n, 0, 15), 1),
  Rating     = sample(1:5, n, replace = TRUE)
)

df$Salary <- 30 +
  2 * df$Years +
  4 * df$Rating +
  6  * (df$Department == "Marketing") +
  15 * (df$Department == "Engineering") +
  rnorm(n, mean = 0, sd = 5)

df$Department <- factor(df$Department,
                        levels=c("HR", "Marketing", "Engineering"))
```

```{r}
model <- lm(Salary ~ Years + Rating + Department, data=df)
summary(model)
```
